Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-15 main.tex
Better explanation for the PoM-Tagger (close #6)

\section{The Translator}
The translator analyzes a parse tree to perform translations. For generating a parse tree of \LaTeX{} expressions, the translator uses the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.}. The \gls*{pom}-Tagger is an \texttt{LL}-Parser defined by a context-free grammar in \gls*{bnf}. Each token will be tagged by meta information defined in manually cultivated lexicon files. \gls*{cas} define their own syntax parser. We were able to use \Maple's internal data structure to get a parse tree of the input.

@@

We integrated the defined translation patterns in the library also into the lexicon files of the \gls*{pom}-Tagger. Thus the tagger also tags a token by its translation patterns to other languages. \Cref{tab:sine-lex-example} gives an example of an entry of the lexicon file.

@@

\section{Forward Translations}\label{sec:forward-translation}
An abstract translator class analyzes each node of the parsed tree and delegates them to specialized subtranslators. In this process, an object called \gls*{teo} is built. This \gls*{teo} is needed to rebuild a string representation of the mathematical expression after all nodes of the parsed tree were translated.

The parsed tree generated by the \gls*{pom}-tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions cannot be disambiguated yet. In consequence, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

@@

\Cref{fig:syntax-tree-usecase} visualizes the \gls*{pom-pt} of the Jacobi polynomial example from~\Cref{tab:JacobiP-usecase}. Because of the differences to expression trees, a backward conversion of the \gls*{pom-pt} to a string representation can be difficult, especially for finding necessary or unnecessary parentheses. Therefore we create the \gls*{teo}.

@@

\begin{table}[t!]
\centering
\sloppy
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcolumntype{Y}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}p{#1}}
\begin{tabularx}{\textwidth}{Y{1.55cm} Y{2.5cm} L L}
	\hline
	& \textbf{Node type} & \textbf{Explanation} & \textbf{Example} \\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ has children}} & Sequence & Contains a list of expressions. & $a+b$ is a sequence with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Balanced Expression & Similar to a sequence. But in this case the sequence is wrapped by \texttt{\tbs left} and \texttt{\tbs right} delimiters. & \texttt{\tbs left(}$a+b$\texttt{\tbs right)} is a balanced expression with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Fraction & All kinds of fractions, such as \texttt{\tbs frac}, \texttt{\tbs ifrac}, etc. & \texttt{\tbs ifrac\{a\}\{b\}} is a fraction with two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Binomial & Binomials & \texttt{\tbs binom\{a\}\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Square Root & The square root with one child. & \texttt{\tbs sqrt\{a\}} has one child ($a$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Radical with a specified index & $n$-th root with two children. & \texttt{\tbs sqrt[a]\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Underscore & The underscore `\_' for subscripts. & The sequence $a$\_$b$ has two children ($a$ and `\_'). The underscore itself `\_' has one child ($b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Caret & The caret `\^{}' for superscripts or exponents. Similar to the underscore. & The sequence $a$\^{}$b$ has two children ($a$ and `\^{}'). The caret itself `\^{}' has one child ($b$).\\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ is a leaf}} & \Macro{} & A semantic \LaTeX{} macro & \texttt{\tbs JacobiP}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Generic \LaTeX{} macro & All kinds of \LaTeX{} macros & \texttt{\tbs Rightarrow}, \texttt{\tbs alpha}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Alphanumerical Expressions & Letters, numbers and general strings. & Depends on the order of symbols. $ab3$ is alphanumerical, while $4b$ are two nodes ($4$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Symbols & All kind of symbols & `$@$', `$*$', `$+$', `$!$', etc.\\
	\hline
\end{tabularx}
\caption{A table of all kinds of nodes in a PoM syntax tree. Note that this table groups some kinds for a better overview. For a complete list and a more detailed version see~\parencite{POM-Tagger}.}
\label{tab:allTypesTable}
\end{table} 
This approach seems to work for most semantic \LaTeX{} expressions. However, there is one problem with this straightforward approach which was mentioned in~\Cref{sec:problems}. There are many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the simple expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between its operands. This only makes sense as long as the operator is a binary operator.

@@

As you can see, the input expression in semantic \LaTeX{} is interpreted as
\begin{equation}\label{eq:multi-expo}
\left( \cos(x)^2 \right)^2.
\end{equation}
 \section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
Instead of writing a custom \Maple{} syntax parser, we use \Maple's internal data structure to get a syntax tree of the input\footnote{In consequence, a license of \Maple{} is mandatory to perform backward translations. The translator is using the version \Maple{} 2016.}. \Maple{} allows several different input styles. The \texttt{1D} input is mainly used for programming purposes and therefore also used to perform our translations. Internally, \Maple{} uses a \gls*{dag} for syntax trees.

>>>>

\section{The Translator}
The translator analyzes a parse tree to perform translations. For generating a parse tree of \LaTeX{} expressions, the translator uses the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.}. \gls*{cas} define their own syntax parser. We were able to use \Maple's internal data structure to get a parse tree of the input. \Cref{sec:forward-translation} and \Cref{sec:backward-translation} will explain the parsing and translation process in detail.

@@

@@

\section{Forward Translations}\label{sec:forward-translation}
As a pre-processing step, we use the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.} for parsing semantic \LaTeX{} expressions. The \gls*{pom}-Tagger is defined by a context-free grammar in \gls*{bnf} and is an \texttt{LL}-Parser, i.e., it parses the input from \textbf{L}eft to right and assigns the \textbf{L}eftmost (first applicable) drivation rule defined by the grammar to an expression. In other words, the \gls*{pom}-Tagger scans the input for \textit{terms} and groups them into subexpressions if suitable, where \textit{terms} are non-terminal symbols in the context of \gls*{bnf}. If a node in the generated parse tree matches a pre-defined symbol, it will be tagged by mata information that are defined in manually cultivated lexicon files.

We integrated the defined translation patterns from our library also into these lexicon files. Thus the tagger also tags a node in the parse tree by its translation patterns. \Cref{tab:sine-lex-example} gives an example of an entry of the lexicon file.

The parsed tree generated by the \gls*{pom}-Tagger is not a mathematical expression tree. The \gls*{pom} project aims to disambiguate mathematical \LaTeX{} expressions and generates an expression tree. However, in the current state, many expressions cannot be disambiguated yet. In consequence, the \gls*{pom}-tagger generates a raw parsed tree where each token in the \LaTeX{} expression is a node in the tree. We call this parsed tree the \gls*{pom-pt}.

@@

\Cref{fig:syntax-tree-usecase} visualizes the \gls*{pom-pt} of the Jacobi polynomial example from~\Cref{tab:JacobiP-usecase}. Because of the differences to expression trees, a backward conversion of the \gls*{pom-pt} to a string representation can be difficult, especially for finding necessary or unnecessary parentheses. Therefore we create the \gls*{teo}. The \gls*{teo} is a list containing already translated subexpressions.

@@

\begin{table}[t!]
\centering
\sloppy
\newcolumntype{L}{>{\raggedright\arraybackslash}X}
\newcolumntype{Y}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}p{#1}}
\begin{tabularx}{\textwidth}{Y{1.55cm} Y{2.5cm} L L}
	\hline
	& \textbf{Node type} & \textbf{Explanation} & \textbf{Example} \\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ has children}} & Sequence & Contains a list of expressions. & $a+b$ is a sequence with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Balanced Expression & Similar to a sequence. But in this case the sequence is wrapped by \texttt{\tbs left} and \texttt{\tbs right} delimiters. & \texttt{\tbs left(}$a+b$\texttt{\tbs right)} is a balanced expression with three children ($a$, $+$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Fraction & All kinds of fractions, such as \texttt{\tbs frac}, \texttt{\tbs ifrac}, etc. & \texttt{\tbs ifrac\{a\}\{b\}} is a fraction with two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Binomial & Binomials & \texttt{\tbs binom\{a\}\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Square Root & The square root with one child. & \texttt{\tbs sqrt\{a\}} has one child ($a$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Radical with a specified index & $n$-th root with two children. & \texttt{\tbs sqrt[a]\{b\}} has two children ($a$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Underscore & The underscore `\_' for subscripts. & The sequence $a$\_$b$ has two children ($a$ and `\_'). The underscore itself `\_' has one child ($b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Caret & The caret `\^{}' for superscripts or exponents. Similar to the underscore. & The sequence $a$\^{}$b$ has two children ($a$ and `\^{}'). The caret itself `\^{}' has one child ($b$).\\
	\hline
	\hline
	\multicolumn{1}{Y{1.5cm}|}{\textbf{$r$ is a leaf}} & \Macro{} & A semantic \LaTeX{} macro & \texttt{\tbs JacobiP}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Generic \LaTeX{} macro & All kinds of \LaTeX{} macros & \texttt{\tbs Rightarrow}, \texttt{\tbs alpha}, etc.\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Alphanumerical Expressions & Letters, numbers and general strings. & Depends on the order of symbols. $ab3$ is alphanumerical, while $4b$ are two nodes ($4$ and $b$).\\
	\cline{2-4}
	\multicolumn{1}{l|}{} & Symbols & All kind of symbols & `$@$', `$*$', `$+$', `$!$', etc.\\
	\hline
\end{tabularx}
\caption{A table of all kinds of nodes in a PoM syntax tree. Note that this table groups some types together for a better overview. For a complete list and a more detailed version see~\parencite{POM-Tagger}.}
\label{tab:allTypesTable}
\end{table} 
This approach seems to work for most semantic \LaTeX{} expressions. However, there is one problem with this straightforward approach which was mentioned in~\Cref{sec:problems}. There are many different types of notations used to represent formulae. \Cref{tab:notations} illustrates the simple expression $(a+b)x$ in different notations. The \gls*{pol}\footnote{Also known as \textit{Warsaw Notation} or \textit{prefix notation}} (hereafter called prefix notation) places the operator to the left of/before its operands. The \gls*{rpol}\footnote{Also known as \textit{postfix notation}} (hereafter called postfix notation) does the opposite and places the operator to the right of/after its operands. The infix notation is commonly used in arithmetic and places the operator between its operands. This only makes sense as long as the operator is a binary operator.

@@

As you can see, the input expression in semantic \LaTeX{} is interpreted as
\begin{equation}\label{eq:multi-expo}
\left( \cos(x)^2 \right)^2.
\end{equation}

\section{\Maple{} to Semantic \LaTeX{} Translator}\label{sec:backward-translation}
Instead of writing a custom \Maple{} syntax parser, we use \Maple's internal data structure to get a syntax tree of the input\footnote{In consequence, a license of \Maple{} is mandatory to perform backward translations. The translator is using the version \Maple{} 2016.}. \Maple{} allows several different input styles. The \texttt{1D} input is mainly used for programming purposes and therefore also used to perform our translations. Internally, \Maple{} uses a \gls*{dag} for syntax trees.
